---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data
Assume we're in the correct working directory.
The data to be processed is in activity.zip in 'root'of assignment. Lets unzip and read the data.
We use a helper function to coerce the text into a date...


```{r unzipload, echo=TRUE}
unzip("activity.zip");

setClass('myDate');
setAs("character","myDate", function(from) as.Date(from, format="%Y-%m-%d"));

activities <- read.csv("activity.csv", header=TRUE, sep=",", colClasses = c("integer", "myDate", "integer"));

```


## What is mean total number of steps taken per day?

We could calculate the mean for each day in one step, however we start by determining the total number of steps for each day and plot those in a histogram. 

First, lets see what the data looks like, just the first few rows.

```{r showfirstrows, echo = TRUE}
head(activities);
```

So, the first column is the number of steps; we use that in our calculation of the totals:


```{r sumstepsday, echo=TRUE}
totalsteps <- aggregate(activities[,1], by = list(activities$date), sum);

```

Now we use the total in our historgram...
```{r histogranmofstes, echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2);
ggplot(totalsteps, aes(totalsteps$x)) + 
  geom_histogram() + 
  labs(title="Histogram of total steps per day", x="Number of steps per day");
```

The mean and median of the total steps per day are:
```{r meanandmedian, echo=TRUE}
mean(totalsteps$x, na.rm = TRUE)
median(totalsteps$x, na.rm = TRUE)
```

## What is the average daily activity pattern?

The steps taken are measured at an interval of 5 minutes, we can use this to see how many steps were taken, on avarage, at a particular interval or moment. One expects less steps at night...

```{r averageatinterval, echo=TRUE}
library(scales)
averageatinterval <- aggregate(activities[,1], by = list(activities$interval), mean, na.rm=TRUE);

ggplot(averageatinterval,aes(Group.1, x)) +
  geom_line() + 
  labs(title="Average steps at an specific moment", x="Time", y="Steps taken") + 
  scale_x_continuous(breaks=c(0, 600, 1200, 1800, 2400), 
                     labels=c("00:00, Midnight",  "06:00", "12:00, Noon", "18:00", "24:00"))
```

Determine the interval/moment of the maximum mean number of steps taken...
The number of steps is x, the interval is Group.1

```{r findmaxstesp, echo=TRUE}
themoment  <- averageatinterval[averageatinterval$x == max(averageatinterval$x),]$Group.1

```

The interval at wich the maximum average of steps taken is at `r themoment`.


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
